\MOC{} Server Pages:

### 2007-04-16 paper/paper.tex
svn-r19: Switched to 'lgrind' for code listings.
All figures and listings are now generated in the paper/ sub-dir,
so there is no need to set TEXINPUTS.

Legend of browse plot is fixed.

Added <?! ?> syntax for auto lifting.  This improves a few examples,
although the tag proliferation is unfortunate.

I have a start on unstaged dir browser in PHP.

darcs-hash:20070416222546-e5a07-bfb60b1d0b4b1e906b9fac6983b3d2386205cf03.gz

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,graphicx,color,listings}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\mylst}[3]{%
  \lstinputlisting[#1 label={lst:#2}, %
    caption={\texttt{#2}.  #3}]{#2}}
\newcommand{\lstfile}{\mylst{}}
\newcommand{\lstfloat}{\mylst{float=htb,}}

\title{\MOC{} Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
\conferenceinfo{\MOC{} Workshop}{October 2004, Vancouver}
\CopyrightYear{2004}
\toappear{Submitted to the \MOC{} Workshop (Vancouver,
  2004).\\\textbf{Note from author:} Sorry, the exposition gets
  really rough toward the end: too much hacking and not enough 
  writing!  If the PC agrees that this is a worthwhile 
  idea, I'll be happy to fix the paper \emph{ex post
    facto\ldots}}

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\lstinline{<?like this?>}.  %
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is your first
\MOC{} server page:

\lstinputlisting{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
If a code block raises an exception, the message is sent to the
client's browser in bold-face, and the rest of the page is
aborted.

\def\cd#1{\lstinline{#1}}
\def\id#1{\texttt{#1}}

A very common use of code blocks is to print out (i.e., send to
the browser) the result of evaluating some expression.  The
syntax \cd{<?= e?>} is designated for this task; \cd{e} must
have type \id{string}.  Alternatively, messages may be formatted
with \id{printf} by placing the format string immediately after
the code delimiter: \lstinputlisting{pi.meta}
\begin{quote}
  ........ $\pi$ ÷ 008 is 0.3927
\end{quote}
In a departure from the standard OCaml syntax for \id{printf}, the
arguments are comma-delimited.
%%
One more kind of code delimiter is used for \emph{declarations;}
these are evaluated during the publishing stage:
\lstinputlisting{perm.meta}
\begin{quote}
Permissions on current directory are 0755
\end{quote}

Once published, this output will never change!  The \id{stat}
call is executed only once, not on each request.

To see how all this works, let's consider how a \MOC{} server
page is translated into a proper \MOC{} program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \id{code}
object which is subsequently run on each request (serve stage).
Listing~\ref{lst:trans.meta} shows a sample \MOC{} server page,
and listing~\ref{lst:trans.ml} contains its translation.

\lstfloat{trans.meta}{This file demonstrates the various
  kinds of code blocks.}%
\begin{lstlisting}[language={[Objective]Caml},
  label={lst:trans.ml}, float=htb, 
  caption={\texttt{trans.ml}.  This is the translation of the
    page in listing~\ref{lst:trans.meta}.  Declarations have
    been lifted, and everything else appears within
    \lstinline{.<brackets>.}}]
module Trans = struct
  let lift x = .<x>.
  (* Declarations *)  
  (* More declarations *) 
  let page  = .<fun req puts -> (*@\label{line:page}@*)
    let arg = Request.arg req in 
    (* Code *) ;
    puts ( (* String to be printed *) );
    puts "Regular text.\n";
    Printf.kprintf puts "format string" ( a )  ( b )  ( c ) ;
    let x = (* Expression *) in (*@\label{line:in}@*)
    (* More code *) ;
    puts "Bye!\n";>.
end
\end{lstlisting}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~\ref{line:page} of
listing~\ref{lst:trans.ml}.  It is a function of two arguments:
\cd{req} encapsulates the HTTP request, including the headers
and query arguments; \cd{puts} is a function with type
\cd{string->unit}; it transmits text across the network to the
visitor's browser.  \cd{arg} looks up an
argument in the query string of the request; its type is
\cd{string->string option}.  These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC{} type system
prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\lstinline{puts "\n"} statements).  It automatically follows
code blocks with either a semi-colon or the \lstinline{in}
keyword, as needed.

Now that we understand how the server page is assembled into a
staged program, the effects of adding \MOC{} staging operators
to our pages should be predictable.  Below is a reformulation of
the permissions example, using escape \cd{.~} and \cd{lift} to
ensure that \id{stat} is called only once.
\lstinputlisting{perm2.meta}
%%
Still, the \id{printf} conversion is performed at serve time, so
maybe it is best to pre-generate that as well:
\lstinputlisting{perm3.meta}
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  Finally it is possible to splice in
additional code at publish time, such as in this count-down
example:
\lstinputlisting{count.meta}
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

@@

\begin{tabular}{l@{\quad$\leadsto$\quad}l}
\cd{<?~ a ?>} &
\cd{<? .~( a ) ?>}
\\
\cd{<?~= b ?>} &
\cd{<?= .~( b ) ?>}
\\
\cd{<?~let x = c ?>} &
\cd{<?let x = .~( c ) ?>}
\\
\cd{<?~"fmt" d, e, f?>} &
\cd{<?"fmt" .~(d), .~(e), .~(f)?>}
\end{tabular}

@@

One last feature in the design of \MOC{} server pages is the ability
to take publish-stage arguments.  The parameter names are specified to
the translator in a \emph{pragma} declaration, which can appear
anywhere in the code:
\begin{lstlisting}[numbers=none]
  <?pragma args x y z ?>
\end{lstlisting}
The translator inserts these as page parameters, like this:
\begin{lstlisting}[numbers=none, language={[Objective]Caml}]
  let page x y z = .< ... >.
\end{lstlisting}
Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC{} code is, for now, an implementation detail
(see \S\ref{sec:impl}).

@@

To explore the expressiveness of this design, we now look at a
series of web services, organized to comprise a small web
site.\footnote{I hope to make the demo available on the web in time
  for the workshop; this will require convincing the I.T. folks to
  open up a non-standard port through the firewall.  For now, it may
  be downloaded and run locally.} %
To give the services a similar look, we developed a site-wide style
sheet, and an OCaml function to generate a navigation bar from a
hierarchical list of page titles and links.  The following bindings are
available site-wide:
\begin{lstlisting}[numbers=none]
  val preamble: string -> string
  val navbar: string -> string
  val postamble: string
\end{lstlisting}
The first function generates the standard XHTML header given the
title, while the second function generates the navigation bar given
the URI denoting the current page.  These can be called at either
stage.  Naturally, if the title includes a dynamic argument, then
\cd{preamble} will be delayed until the serve stage.

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been pre-generated.  It is
not possible, with the current implementation, for the user to request
other exponents (see \S\ref{sec:impl} for an explanation).  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \id{Num} module.  The implementation of this
service appears in listing~\ref{lst:power.meta}.

@@

\lstfile{power.meta}{This implements the staged power function as a
  web service.  The exponent \cd{y} is an early (publish stage)
  argument; but the base \cd{x} is late (serve stage).}

@@

\mylst{lastline=39,}{dir.meta}{Implementation of directory
  browsing service.  This is the page part; some top-level
  declarations were omitted.  See also
  listing~\ref{lst:dir.meta-2}.}

We omitted some of the more mundane functions on this page, but
\cd{list_files} is important, so it appears in
listing~\ref{lst:dir.meta-2}.  In the first stage, this reads
all the filenames in the directory and sorts them by name.  Then
it looks up the fileinfo, computes the MD5 sum, and prepares the
text of each row.  Finally, it generates code that matches the
filename against a (future-stage) regular expression, and
includes it in a list only if it matches.  In the next stage,
this list will be constructed and sorted (if necessary).

\mylst{}{dir.meta-2}{This top-level function transforms a list
  of filenames into a code sequence that tests each name against
  a regexp, and adds it to the list only in the case of a
  match.}

@@

\mylst{lastline=30,}{gc.meta}{...}

@@

Currently, the \MOC{} implementation is not capable of reading
and writing \id{code} values to a file.  This means that
generated server pages (after the publish stage) must already be
in the server memory when it starts.  This is an unfortunate
limitation, as it means we must restart the entire web server to
re-publish any particular page!

A generated server page has this type:
\begin{lstlisting}[numbers=none]
  type code = Request.req -> (string -> unit) -> unit
\end{lstlisting}
As the server starts, it constructs a dictionary (\cd{code
  StringMap.t}) mapping URIs to code.  This dictionary is
consulted on each HTTP request.  If the URI is not found there,
then we check the file system, in case it refers to a static
file.  If this fails, we respond with a 404 (resource not
found).

@@

Figure~\ref{fig:static} includes comparable results for PHP, the
popular server-side computation system.\footnote{libphp4.so
  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
static data files the extension \texttt{.php}, so that Apache
would treat them as PHP code, even though they have no
\lstinline{<?code blocks?>}.
%%
We omitted the results for Apache serving static files, because they
are way off scale: Apache handled an astounding 4,438 hits per second
for the 1k file, and 1,813 for the 64k file.  Apache is heavily
optimized for serving static files: apart from caching, it uses the
special \texttt{sendfile()} system call for zero-copy file transfer
from kernel space~\cite{tranter03sendfile}.  Although our prototype is
no match for Apache on static files, the overhead for interpreting
code seems no worse than that of PHP.
%%

@@

\begin{itemize}

\item This really needs to be a module for a real server, such
  as Apache or AOLserver.

\item Need more serious case studies, where one logical service
  is implemented across a series of pages.

\item Need to address the major implementation shortcoming, of
  loading all the generated pages into memory on startup.

@@

\end{itemize}

>>>>

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,graphicx,color}
\usepackage[leftno,lineno5]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

\newcounter{listing}
\setcounter{listing}{0}
\makeatletter
\def\ext@listing{lol}
\def\fnum@listing{Listing\nobreakspace\thelisting}
\makeatother
\newcommand{\mylst}[2]{%
  \captionof{listing}{\textbf{#1.} #2}%
  \label{lst:#1}%
  \lgrindfile{#1}}

\title{\MOC{} Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
\conferenceinfo{\MOC{} Workshop}{October 2004, Vancouver}
\CopyrightYear{2004}
\toappear{Submitted to the \MOC{} Workshop (Vancouver,
  2004).}

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\cd{\Mquote{}like this\Munquote}.
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is your first
\MOC{} server page:
\lgrindfile{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
If a code block raises an exception, the message is sent to the
client's browser in bold-face, and the rest of the page is
aborted.

A very common use of code blocks is to print out (i.e., send to
the browser) the result of evaluating some expression.  The
syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
have type \cd{string}.  Alternatively, messages may be formatted
with \cd{printf} by placing the format string immediately after
the code delimiter: \lgrindfile{pi.meta}
\begin{quote}
  ........ $\pi$ ÷ 008 is 0.3927
\end{quote}
In a departure from the standard OCaml syntax for \cd{printf}, the
arguments are comma-delimited.
%%
One more kind of code delimiter is used for \emph{declarations;}
these are evaluated during the publishing stage:
\lgrindfile{perm.meta}
\begin{quote}
Permissions on current directory are 0755
\end{quote}

Once published, this output will never change!  The \cd{stat}
call is executed only once, not on each request.

To see how all this works, let's consider how a \MOC{} server
page is translated into a proper \MOC{} program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
Listing~\ref{lst:trans.meta} shows a sample \MOC{} server page,
and listing~\ref{lst:trans.ml} contains its translation.

\mylst{trans.meta}{This file demonstrates the various
  kinds of code blocks.}

\mylst{trans.ml}{This is the translation of the
    page in figure~\ref{lst:trans.meta}.  Declarations have
    been lifted, and everything else appears within
    \cd{\Mbegincode brackets\Mendcode}}

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~\ref{line:page} of
listing~\ref{lst:trans.ml}.  It is a function of two arguments:
\cd{req} encapsulates the HTTP request, including the headers
and query arguments; \cd{puts} is a function with type
\cd{string\to unit}; it transmits text across the network to the
visitor's browser.  \cd{arg} looks up an
argument in the query string of the request; its type is
\cd{string\to string option}.  These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC{} type system
prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically follows
code blocks with either a semi-colon or the {\KWfont in}
keyword, as needed.

Now that we understand how the server page is assembled into a
staged program, the effects of adding \MOC{} staging operators
to our pages should be predictable.  Below is a reformulation of
the permissions example, using escape \Msplice{} and \cd{lift} to
ensure that \cd{stat} is called only once.
\lgrindfile{perm2.meta}
%%
Still, the \cd{printf} conversion is performed at serve time, so
maybe it is best to pre-generate that as well:
\lgrindfile{perm3.meta}
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  Finally it is possible to splice in
additional code at publish time, such as in this count-down
example:
\lgrindfile{count.meta}
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

@@

\begin{tabular}{l@{\quad$\leadsto$\quad}l}
\cd{\Mquote{\Mtilde} a \Munquote} &
\cd{\Mquote{} \Msplice( a ) \Munquote}
\\
\cd{\Mquote{\Mtilde=} b \Munquote} &
\cd{\Mquote{=} \Msplice( b ) \Munquote}
\\
\cd{\Mquote{\Mtilde}let x = c \Munquote} &
\cd{\Mquote{} let x = \Msplice( c ) \Munquote}
\\
\cd{\Mquote{\Mtilde}\id{"fmt"} d, e, f\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(d), \Msplice(e), \Msplice(f)\Munquote}
\\
\end{tabular}

@@

One last feature in the design of \MOC{} server pages is the ability
to take publish-stage arguments.  The parameter names are specified to
the translator in a \emph{pragma} declaration, which can appear
anywhere in the code:

(EXPLAIN THIS INSTEAD USING TRANS FIGURE)

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC{} code is, for now, an implementation detail
(see \S\ref{sec:impl}).

@@

To explore the expressiveness of this design, we now look at a
series of web services, organized to comprise a small web
site.
%\footnote{I hope to make the demo available on the web in time
%  for the workshop; this will require convincing the I.T. folks to
%  open up a non-standard port through the firewall.  For now, it may
%  be downloaded and run locally.} %
To give the services a similar look, we developed a site-wide style
sheet, and an OCaml function to generate a navigation bar from a
hierarchical list of page titles and links.  The following bindings are
available site-wide:

\lgrindfile{navspec-sig.ml}

The first function generates the standard XHTML header given the
title, while the second function generates the navigation bar given
the URI denoting the current page.  These can be called at either
stage.  Naturally, if the title includes a dynamic argument, then
\cd{preamble} will be delayed until the serve stage.

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been pre-generated.  It is
not possible, with the current implementation, for the user to request
other exponents (see \S\ref{sec:impl} for an explanation).  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The implementation of this
service appears in listing~\ref{lst:power.meta}.

@@

\mylst{power.meta}{This implements the staged power function as a
  web service.  The exponent \cd{y} is an early (publish stage)
  argument; but the base \cd{x} is late (serve stage).}

@@

\mylst{dir.meta}{...}

In the first stage, this reads
all the filenames in the directory and sorts them by name.  Then
it looks up the fileinfo, computes the MD5 sum, and prepares the
text of each row.  Finally, it generates code that matches the
filename against a (future-stage) regular expression, and
includes it in a list only if it matches.  In the next stage,
this list will be constructed and sorted (if necessary).

@@

%\lgrindfile{gc.meta}

@@

Currently, the \MOC{} implementation is not capable of reading
and writing \cd{code} values to a file.  This means that
generated server pages (after the publish stage) must already be
in the server memory when it starts.  This is an unfortunate
limitation, as it means we must restart the entire web server to
re-publish any particular page!

A generated server page has this type:

\lgrindfile{codeHandler.mli}

As the server starts, it constructs a dictionary (\cd{code
  StringMap.t}) mapping URIs to code.  This dictionary is
consulted on each HTTP request.  If the URI is not found there,
then we check the file system, in case it refers to a static
file.  If this fails, we respond with a 404 (resource not
found).

@@

Figure~\ref{fig:static} includes comparable results for PHP, the
popular server-side computation system.\footnote{libphp4.so
  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
static data files the extension \texttt{.php}, so that Apache
would treat them as PHP code, even though they have no
\cd{\Mquote{}code blocks\Munquote}.
%%
We omitted the results for Apache serving static files, because they
are way off scale: Apache handled an astounding 4,438 hits per second
for the 1k file, and 1,813 for the 64k file.  Apache is heavily
optimized for serving static files: apart from caching, it uses the
special \texttt{sendfile()} system call for zero-copy file transfer
from kernel space~\cite{tranter03sendfile}.  Although our prototype is
no match for Apache on static files, the overhead for interpreting
code seems no worse than that of PHP.
%%

@@

 This really needs to be a module for a real server, such
  as Apache or AOLserver.

 Need more serious case studies, where one logical service
  is implemented across a series of pages.

 Need to address the major implementation shortcoming, of
  loading all the generated pages into memory on startup.

@@
